Co-operation between the sexes trumps selfish control of female meiotic drive. Or Sperm do not evolve to collaborate in female meiotic drive. Or Why does female meiosis allow the opportunity for exploitation by selfish sperm?

### 2014-02-16 writeup_and_notes/male_dependent_drive.tex
adding to appendix

%Gametogenesis and meiosis are common times for genetic conflict. 
%This is a consequence of the fact that as the set of alleles that consitute
%	an individual are in the process of parting ways, 
%	with only a subset of them being assured transmission.
%In constrast, the perspective of the fertilizing sperm's genome is different, it has to live with the consequences of any distortion it causes 
%to female meiosis through the entire next life cycle of the organism. 
%It is therefore in the interest of the fertilizing sperm's genome to
%make a high fitness zygote even if that means alleles in sperm
%disrupting other copies of themselves from gaining a 
%transmission advantage through female meiosis.

\section*{Appendix}
In our first scenerio allele $A$ is the ancestral allele and allele
$B$ is our sperm dependent drive allele.
The sperm dependence of the drive is a form of assortative
transmission (mating), and so we have to keep track of the genotype frequencies.
Our genotype frequencies at birth are $x_{AA},~x_{AB},$ and
$x_{BB}$. These three genotypes have fitnesses $w_{AA},~w_{AB},$ and
$w_{BB}$, which are identical between males and females. As the
fitnesses are the same in males and females the frequencies of the
genotypes are identical in males and females, but in the following
equations they are marked with $\Mars$ and $\Venus$ respectively to
allow the reader to more easily follow the logic.
The frequency of the three genotypes at birth in the next generation is:
\begin{equation}
x'_{BB} =\frac{1}{\bar{w} } \Big( w_{BB} x_{BB}^{\Venus} (w_{BB} x_{BB}^{\Mars} +\frac{1}{2}
  w_{AB} x_{AB}^{\Mars} ) + w_{AB} x_{AB}^{\Venus} (d w_{BB}
  x_{BB}^{\Mars} +\frac{1}{2} d
w_{AB} x_{AB}^{\Mars} ) \Big),  \label{eqn.A1}
\end{equation}
\begin{equation}
x'_{AA} =\frac{1}{\bar{w} } \Big(  w_{AA} x_{AA}^{\Venus} (w_{AA} x_{AA}^{\Mars} + \frac{1}{2}
   w_{AB} x_{AB}^{\Mars}) + w_{AB} x_{AB}^{\Venus}(\frac{1}{2} w_{AA} x_{AA}^{\Mars}+ \frac{1}{2} 
  w_{AB} x_{AB}^{\Mars}) \Big),  \label{eqn.A2}
\end{equation}
and
\begin{eqnarray}
x'_{AB} = & \frac{1}{\bar{w} } \Big( w_{AA} x_{AA}^{\Venus}  ( w_{BB}  x_{BB}^{\Mars} + \frac{1}{2}
  w_{AB}  x_{AB}^{\Mars} ) +w_{BB} x_{BB}^{\Venus} ( w_{AA}  x_{AA}^{\Mars} + \frac{1}{2}
  w_{AB}  x_{AB}^{\Mars} )  \nonumber \\
& + x_{AB}^{\Venus}(\frac{1}{2}  w_{AA}  x_{AA}^{\Mars}+\frac{1}{2} (1-d)
 w_{AB}  x_{AB}^{\Mars} + \frac{1}{4}  w_{AB}
 x_{AB}^{\Mars} + (1-d)  w_{BB}  x_{BB}^{\Mars} ) \Big) \label{eqn.A3}
\end{eqnarray}
where the mean fitness of the population is
\begin{equation}
\bar{w} = w_{AA} x_{AA}+w_{AB}  x_{AB} +w_{BB}  x_{BB}.
\end{equation}
We iteration can then iterate this recursion over generations, for
example to produce the frequency trajectories shown in the inset
Figure \ref{Invasion_space} by plotting $x_{BB}+ \frac{1}{2}x_{AB}$
over time. We can also use this to assess the
invasion criteria for the allele by starting allele B at very low
frequency, and seeing if it increases across the generations.

\paragraph{Dominance} In the supplement we also explore a model where
the genotype of the male who produced the fertilizing sperm affects female drive,
instead of just the allelic type of the sperm. We assume that if a
heterozygous egg ($AB$) is fertilized by a spermwith the following
male ($\Mars$) genotypes the $B$ allele will be transmitted through
female meiosis with the follow probabilites
\begin{tabular}{cccc}
$\Mars$ geno. & AA & AB  & BB\\
B transmission prob. & $\frac{1}{2}$ &  $\frac{1}{2} + h(d-\frac{1}{2})$  & $d$\\ 
\end{tabular}
where $d > 1/2$ and $h$ ($0<h<1$) acts to specify the dominance of the $B$
allele's effect on female meiotic drive through the males genotype. We
set up a set of recursions, similar to those in eqn.
\ref{eqn.A1}-\ref{eqn.A1}, but now featuring these male genotype
transmission probabilites. In Supplementary Figure \ref{Effect_of_dominance} we show the invasion
criteria for both the A and the B allele when they are rare in the
population. We find that there is a narrow parameter range the B allele
can invade but not fix if it is dominant (i.e. $h>0$) and the cost in homozygotes
is not too great. This reflects the fact that when the allele is rare
it can increase in frequency due to fertilization of heterozygous eggs
by sperm from heterozygote males, where the sperm do not themselves
carry the allele. But as the B allele becomes common it will frequently be
present in the homozygous state in males, and so will be unable to
avoid the production of unfit homozygote zygotes leaving it trapped in
the population.

@@

>>>>

%Gametogenesis and meiosis are common times for genetic conflict. 
%This is a consequence of the fact that as the set of alleles that consitute
%	an individual are in the process of parting ways, 
%	with only a subset of them being assured transmission.
%In constrast, the perspective of the fertilizing sperm's genome is different, it has to live with the consequences of any distortion it causes 
%to female meiosis through the entire next life cycle of the organism. 
%It is therefore in the interest of the fertilizing sperm's genome to
%make a high fitness zygote even if that means alleles in sperm
%disrupting other copies of themselves from gaining a 
%transmission advantage through female meiosis.
  \appendix
\section*{Appendix}
In our first scenerio allele $A$ is the ancestral allele and allele
$B$ is our sperm dependent drive allele.
The sperm dependence of the drive is a form of assortative
transmission (mating), and so we have to keep track of the genotype frequencies.
Our genotype frequencies at birth are $x_{AA},~x_{AB},$ and
$x_{BB}$. These three genotypes have fitnesses $w_{AA},~w_{AB},$ and
$w_{BB}$, which are identical between males and females. As the
fitnesses are the same in males and females the frequencies of the
genotypes are identical in males and females, but in the following
equations they are marked with $\Mars$ and $\Venus$ respectively to
allow the reader to more easily follow the logic.
The frequency of the three genotypes at birth in the next generation is:
\begin{equation}
x'_{BB} =\frac{1}{\bar{w} } \Big( w_{BB} x_{BB}^{\Venus} (w_{BB} x_{BB}^{\Mars} +\frac{1}{2}
  w_{AB} x_{AB}^{\Mars} ) + w_{AB} x_{AB}^{\Venus} (d w_{BB}
  x_{BB}^{\Mars} +\frac{1}{2} d
w_{AB} x_{AB}^{\Mars} ) \Big),  \label{eqn.A1}
\end{equation}
\begin{equation}
x'_{AA} =\frac{1}{\bar{w} } \Big(  w_{AA} x_{AA}^{\Venus} (w_{AA} x_{AA}^{\Mars} + \frac{1}{2}
   w_{AB} x_{AB}^{\Mars}) + w_{AB} x_{AB}^{\Venus}(\frac{1}{2} w_{AA} x_{AA}^{\Mars}+ \frac{1}{2} 
  w_{AB} x_{AB}^{\Mars}) \Big),  \label{eqn.A2}
\end{equation}
and
\begin{eqnarray}
x'_{AB} = & \frac{1}{\bar{w} } \Big( w_{AA} x_{AA}^{\Venus}  ( w_{BB}  x_{BB}^{\Mars} + \frac{1}{2}
  w_{AB}  x_{AB}^{\Mars} ) +w_{BB} x_{BB}^{\Venus} ( w_{AA}  x_{AA}^{\Mars} + \frac{1}{2}
  w_{AB}  x_{AB}^{\Mars} )  \nonumber \\
& + x_{AB}^{\Venus}(\frac{1}{2}  w_{AA}  x_{AA}^{\Mars}+\frac{1}{2} (1-d)
 w_{AB}  x_{AB}^{\Mars} + \frac{1}{4}  w_{AB}
 x_{AB}^{\Mars} + (1-d)  w_{BB}  x_{BB}^{\Mars} ) \Big) \label{eqn.A3}
\end{eqnarray}
where the mean fitness of the population is
\begin{equation}
\bar{w} = w_{AA} x_{AA}+w_{AB}  x_{AB} +w_{BB}  x_{BB}.
\end{equation}
We iteration can then iterate this recursion over generations, for
example to produce the frequency trajectories shown in the inset
Figure \ref{Invasion_space} by plotting $x_{BB}+ \frac{1}{2}x_{AB}$
over time. We can also use this to assess the
invasion criteria for the allele by starting allele B at very low
frequency, and seeing if it increases across the generations.

\paragraph{cost of driver in heterozygotes}
We also explored a range of other costs to the drive alleles other than a
simple recessive cost, for example an additive cost. We modified the recursions given in eqn
\ref{eqn.A1}-\ref{eqn.A3} to feature an additive cost $s$ for each
copy of the $B$ allele. A none recessive cost to the drive allele can
cause bistability of the drive allele. The B allele is unable to enter
the population when rare, this results from the fact that
when the allele is rare ($x_B \ll 1$) sperm carrying it rarely
encounter heterozygous eggs ($\sim x^2$) but the allele still pays a
cost in heterozygous individuals.  While when the B allele is common ($x_B
\gg 0$) encounters between heterozygous eggs and $B$ allele sperm
become more common allowing drive to occur this can allow the $B$
allele to spread when common. In Supplementary Figure
\ref{bistable_additive} we show the unstable equilibrium frequency
above which the $B$ allele can invade. This equilibrium frequency was found by a simple
parameter search.  \gc{wonder if a bit of work could find this}

\paragraph{Dominance} In the supplement we also explore a model where
the genotype of the male who produced the fertilizing sperm affects female drive,
instead of just the allelic type of the sperm. We assume that if a
heterozygous egg ($AB$) is fertilized by a sperm with the following
male ($\Mars$) genotypes the $B$ allele will be transmitted through
female meiosis with the follow probabilites\\
\begin{tabular}{cccc}
$\Mars$ geno. & AA & AB  & BB\\
P(B). & $\frac{1}{2}$ &  $\frac{1}{2} + h(d-\frac{1}{2}) $ & $d$ 
\end{tabular}\\
where $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
allele's effect on female meiotic drive through the males genotype. We
set up a set of recursions using a recessive sex independent cost to
the driver, similar to those in eqn.
\ref{eqn.A1}-\ref{eqn.A1}, but now featuring these male genotype
transmission probabilites. In Supplementary Figure \ref{Effect_of_dominance} we show the invasion
criteria for both the A and the B allele when they are rare in the
population. We find that there is a narrow parameter range the B allele
can invade but not fix if it is dominant (i.e. $h>0$) and the cost in homozygotes
is not too great. This reflects the fact that when the allele is rare
it can increase in frequency due to fertilization of heterozygous eggs
by sperm from heterozygote males, where the sperm do not themselves
carry the allele. But as the B allele becomes common it will frequently be
present in the homozygous state in males, and so will be unable to
avoid the production of unfit homozygote zygotes leaving it trapped in
the population. \gc{Perhaps we should show the equil. freq. in the
  fig, as I'm guessing it might be rare}

@@

\gc{Alternatively we could give the table below:}
